---
title: "poisson_regression"
output: html_document
date: "2024-11-08"
---

## Load Necessary Libraries

```{r}
library(haven) # Needed for reading in .dat files
library(tidyverse) 
library(datasets)
library(assertr) # For data checks
library(eeptools) # For data checks

```

## Read in Ship Damage Incidents Dataset

### Variables Summary :

#### type : Factor with levels "A" to "E" for the different ship types

#### construction : factor with levels "1960-64", "1965-69", "1970-74", "1975-79" for the periods of ship construction

#### operation : Factor with levels "1960-74", "1975-79" for the periods of operation

#### months : Aggregate months of service, for all ships of specified type, construction, and operation

#### damage : Number of incidents of wave damage for cargo carrying ships

```{r}
ship_data <- read.table("/Users/fillm/Desktop/Job Search Materials/Materials for Portfolio Project Creation/Poisson Regression/dataset/ships.dat", header = TRUE)
```

## Perform Data Checks

```{r}
## Confirm that all rows have at least 1 month of exposure
ship_data |> assert(within_bounds(0, Inf, include.lower = FALSE, include.upper = FALSE), "months")

## View unique values of each predictor variable, ensure values are consistent with what we expect them to contain 
unique(ship_data$type)
unique(ship_data$construction)
unique(ship_data$operation)

## Check that rows are unique by predictors
isid(ship_data, vars = c("type", "construction", "operation")) # Returns TRUE, signifying the predictor columns together uniquely identify rows, as expected

```

## Create a Column for the Damage Incident Rate

```{r}
## Damage Incident Rate
## Need to use a rate as the number of service months (exposure) are not equal across rows
ship_data <- ship_data |>
  mutate(
    damage_rate = damage / months
  )

```

## Check Assumption that the Independent Variable (Damage Rate) is poisson distributed

```{r}
## Create a histogram of damage rates, should follow a poisson distribution
hist(ship_data$damage_rate) # Appears possibly poisson distributed at first glance

## Check Independent variable mean and variance and compare them
mean(ship_data$damage_rate)
var(ship_data$damage_rate)
mean(ship_data$damage_rate) / var(ship_data$damage_rate) # Mean and variance differ, suggesting equidispersion is not met 

## Run a Kolmogorov-Smirnov Test to determine if damage rate is poisson distributed
estimated_mean <- mean(ship_data$damage_rate) #use the mean of our ship dataset to use as the mean of poisson distribution we will test it against
ks.test(ship_data$damage_rate, "ppois", lambda = estimated_mean) # extremely low p-value, reject null and conclude data are not poisson distributed

```

## Create A Model of Damage Incident Rate

### Use Quasi-Poisson as we are modeling count data, and the mean-variance condition appears to not be met

### Including Ship type (type), Construction Period (construction), and Operation Period (operation) as predictors

### Including Number of Months (months) as an offset term to control for differing exposures in each row

```{r}
## Create Quasi-Poisson Regression Model
ship_model <- glm(damage ~ type + construction + operation + offset(log(months)), family = quasipoisson(link = "log"), data = ship_data)

## Output a Summary of the Model
summary(ship_model)

```

## Comment on which coefficients are statistically significant per model summary above :

#### - Coefficient "typeB" is significant at a .05 alpha level

#### - Coefficients "construction1965-69" and "construction1970-74" are both significant at a .01 alpha level

#### - Coefficient "operation1975-79" is significant at a .05 alpha level

#### - None of the remaining coefficients were found to be statistically significant

## Interpret Statistically Significant Model Coefficients in Context

### Note : Model base group are ships of type=A, construction="1960-64", operation="1960-74"

### Exponentiate the Coefficients exp(coeff) to obtain risk as compared to Base Group

#### Increasing predictor variable by 1 unit has a multiplicative effect of exp(coeff) on the predicted mean damage count

##### - typeB : Keeping all other variables constant, ships of Type B are predicted by the model to have 0.58 times the number of incidents of damage as compared to the base group

##### - construction1965-69 : Keeping all other variables constant, ships constructed in 1965-1969 are predicted by the model to have 2.01 times the number of incidents of damage as compared to the base group

##### - construction1970-74 : Keeping all other variables constant, ships constructed in 1970-1974 are predicted by the model to have 2.27 times the number of incidents of damage as compared to the base group

##### - operation1975-79 : Keeping all other variables constant, ships operating in 1975-1979 are predicted by the model to have 1.47 times the number of incidents of damage as compared to the base group

```{r}
## Exponentiating Coefficients and rounding to two decimal places
round(exp(coef(ship_model)), 2)
```
